---
title: "Wine Quality Logistic Regression"
author: "Jim Harner"
date: "7/25/2018"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

library(glmnet)
library(dplyr)
library(sparklyr)
sc <- spark_connect(master = "local")
```

## 7.5 Wine Quality Logistic Regression

This section explores feature selection based on regularization.

### 7.5.1 Regularized Logistic Regression with Spark

We now revisit the Wine Quality Data Set analyzed in Section 7.3.1. Our goal is to continue with advanced analyses.

We read the `winequality-red.csv` file into a Spark DataFrame using `spark_red_csv`. We want to ensure the training and test data frames are identical to those in Section 7.3.1.
```{r}
wine_red_sdf <- spark_read_csv(sc, "wine_red_sdf",
    path = "file:///home/rstudio/rspark-tutorial/data/wine/winequality-red.csv",
    delimiter = ";" )
wine_red_tbl <- sdf_register(wine_red_sdf, name = "wine_red_tbl")
```

We split `wine_red_sdf` into a training and a test Spark DataFrame as before and cast `quality` as numeric in order to binarize it with a threshold.
```{r}
wine_red_partition <- wine_red_tbl %>%
  mutate(quality = as.numeric(quality)) %>%
  ft_binarizer(input.col = "quality", output.col = "quality_bin",
               threshold = 5.0) %>%
  sdf_partition(training = 0.7, test = 0.3, seed = 2)
wine_red_train_sdf <- wine_red_partition$training
wine_red_test_sdf <- wine_red_partition$test
```

The full model is now run.
```{r}
wine_red_br_full_fit <- wine_red_train_sdf %>%
  ml_logistic_regression(quality_bin ~ fixed_acidity + volatile_acidity
                         + citric_acid + residual_sugar + chlorides
                         + free_sulfur_dioxide + total_sulfur_dioxide
                         + density + pH + sulphates + alcohol)
summary(wine_red_br_full_fit)
```

```{r}
wine_red_br_fnc <- function(l) {
  wine_red_train_sdf %>%
    ml_logistic_regression(quality_bin ~ fixed_acidity + volatile_acidity
                           + citric_acid + residual_sugar + chlorides
                           + free_sulfur_dioxide + total_sulfur_dioxide
                           + density + pH + sulphates + alcohol,
                           elastic_net_param = 1, reg_param = l)
}
```

```{r}
regParm <- c(0.0, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4)
wine_red_br_auc <- data.frame(lambda = 0,
                              auc = wine_red_br_full_fit$summary$area_under_roc)
wine_red_br_coef <- as.data.frame(wine_red_br_full_fit$coefficients)
```

We now calculate the `coefficients` and `auc` for each of the models.
```{r}
for(l in regParm) {
  wine_red_br_fit <- wine_red_br_fnc(l)
  wine_red_br_auc <- data.frame(lambda = l,
                        auc = wine_red_br_fit$summary$area_under_roc) %>%
    rbind(wine_red_br_auc, .)
  wine_red_br_coef <- 
    as.data.frame(wine_red_br_fit$coefficients) %>%
    cbind(wine_red_br_coef, .)
}
wine_red_br_auc
```
```{r}
library(ggplot2)
wine_red_br_auc %>%
  ggplot(aes(x = lambda)) +
  geom_point(aes(y = auc, color = 'auc')) +
  geom_line(aes(y = auc, color = 'auc')) +
  ggtitle("Performance Metric for the Red Wine Regulated Models") +
  xlab("Lambda") + ylab("AUC")
```

```{r}
names(wine_red_br_coef) <- as.character(rbind(c(0.0, regParm)))
wine_red_br_coef <- t(wine_red_br_coef)
wine_red_br_coef
```

The interpretation is better if we visualize the coefficient traces.
```{r}
library(ggplot2)
as.data.frame(cbind(lambda = c(0.0, regParm), wine_red_br_coef)) %>%
  ggplot(aes(x = lambda)) +
  geom_line(aes(y = fixed_acidity, color = 'fixed_acidity')) +
  geom_line(aes(y = volatile_acidity, color = 'volatile_acidity')) + 
  geom_line(aes(y = citric_acid, color = 'citric_acid')) + 
  geom_line(aes(y = residual_sugar, color = 'residual_sugar')) + 
  geom_line(aes(y = chlorides, color = 'chlorides')) + 
  geom_line(aes(y = free_sulfur_dioxide, color = 'free_sulfur_dioxide')) + 
  geom_line(aes(y = total_sulfur_dioxide, color = 'total_sulfur_dioxide')) +
  geom_line(aes(y = density, color = 'density')) +
  geom_line(aes(y = pH, color = 'pH')) +
  geom_line(aes(y = sulphates, color = 'sulphates')) +
  geom_line(aes(y = alcohol, color = 'alcohol')) +
  ggtitle("Parameter Trace for the Red Wine Regulated Models") +
  xlab("Lambda") + ylab("Estimate")
```

The coefficients go to 0 very quickly. Based on regularization, `alcohol` and `density` are still standing at $\lambda = 0.2$, but then they too go to 0. Based on feature importance, `alcohol`, was instead most important.

We now `collect` the training and test Spark DataFrames into R as regular data frames. If you experiment with `alpha` and `lambda`,i.e., invoke the elastic net, you will see the coefficients that are driven to 0 vary greatly. 
```{r}
wine_red_train_df <- collect(wine_red_partition$training)
wine_red_test_df <- collect(wine_red_partition$test)
```

### 7.5.2 Regularized Logistic Regression with glmnet

We can now use `glmnet` to model the wine quality.
```{r}
wine_red.x <- model.matrix(as.factor(quality_bin) ~ fixed_acidity 
                           + volatile_acidity + citric_acid + residual_sugar
                           + chlorides + free_sulfur_dioxide
                           + total_sulfur_dioxide + density + pH + sulphates
                           + alcohol, 
                           data = wine_red_train_df)[, -1]
wine_red.y <- wine_red_train_df$quality_bin

wine_red_bin <- glmnet(x = wine_red.x, y = wine_red.y, family = "binomial",
                       alpha = 1, lambda = c(0.0, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4),
                       standardize = TRUE)
coef(wine_red_bin, s = c(0.0, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4))
```
You can experiment with different values of `alpha` and `lambda`.

```{r}
spark_disconnect(sc)
```

